Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-16 treestats_paper.tex
Review comments from Peter.

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We exploit this duality here to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two (dual) ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
with a large number of mutations under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
bespoke matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may hold interesting biological
signals.

\jk{Not sure if the last bit is expressed the right way}

@@

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarized} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarized)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarized statistics is less tight
than for the unpolished ones.

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialization in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous and tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgments}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
and Leo Speidel for providing the inferred 1000 Genomes tree sequence.

>>>>

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We exploit this duality here to create a method that efficiently computes single-site statistics
from population genetic data using the succinct tree sequence encoding
of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two (dual) ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit) and suggest several new statistics.
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimised matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may hold interesting biological
signals.

@@

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarized} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarized)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarized statistics is less tight
than for the unpolarized ones.

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialization in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous tree and are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgments}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, Kelley Harris and
Aaron Ragsdale for useful suggestions
and Leo Speidel for providing the inferred 1000 Genomes tree sequence.
